---
title: "hw-3 code"
output: html_notebook
---

First i will load library and clean memory

```{r}
#Load Libraries
rm(list = ls()) # clean memory
library(tidyverse)
library(readxl)
library(ggplot2)
library(qicharts2)
library(scales)

```

I will now load my data

```{r}
#Load Data


file_path <- "C:/Users/lgianetti/OneDrive - City of Cambridge/DACSS Program/690V_data_viz/hw-3/ncesdata_MASS_202425.xlsx"

my_data <- read_excel(path = file_path,
    skip = 11)

```

I will do a quick look at the data

```{r}
#Look at Data
str(my_data, width = 70,strict.width='cut')
```

Now I will clean the data by correcting the data types and checking for missing values and outliers. I will also convert the appropriate columns to factors and numeric types as needed.

```{r}
#Clean Data
my_data$`Low Grade` <- factor(my_data$`Low Grade`, ordered = TRUE)
my_data$`High Grade` <- factor(my_data$`High Grade`, ordered = TRUE)
my_data$`Locale Code` <- factor(my_data$`Locale Code`)
my_data$`Locale` <- factor(my_data$`Locale`)
my_data$`Charter` <- factor(my_data$`Charter`)
my_data$`Type` <- factor(my_data$`Type`)
my_data$`Status` <- factor(my_data$`Status`)
my_data$`Students` <- as.numeric(my_data$`Students`)
my_data$`Teachers` <- as.numeric(my_data$`Teachers`)
my_data$`Student Teacher Ratio` <- as.numeric(my_data$`Student Teacher Ratio`)
my_data$`Directly Certified` <- as.numeric(my_data$`Directly Certified`)
my_data$`County Name` <- factor(my_data$`County Name`) 


str(my_data, width = 70,strict.width='cut')
summary(my_data) # check for missing values and outliers



```

Now I will start to explore the data by creating some visualizations. I will create a bar plot to show the distribution of schools by County.

```{r}

# Total values
TotalValues=table(my_data$'County Name',
                 useNA = "ifany") #include all values!
TotalValues

TotalPercent = prop.table(TotalValues)*100
```

```{r}
# Create a data frame for plotting
CountyTable=as.data.frame(TotalValues)

names(CountyTable)=c("County","Count")

CountyTable$Percent=as.vector(TotalPercent)

CountyTable
```

```{r}
# Create a bar plot
base = ggplot(CountyTable,
              aes(x=reorder(County, -Percent),
                  y=Percent))

base + geom_bar(stat="identity")
```

let's explore student teacher ratio. I need to clean the data first. THere are NA associated with missing values. There are 10 rows with missing student and/or teach data and should be removed from the analysis. There are also two extereme outliers with student teacher ratios \>2500 which is not valid and likely due to data entry issue with the teacher value. i will impute these to the mean value for the county.

```{r}


#impute outliers with mean value for the county. Impute any student teacher ratio over 100 with the mean value for the county
my_data <- my_data %>%
  group_by(`County Name`) %>%
  mutate(`Student Teacher Ratio` = ifelse(`Student Teacher Ratio` > 100,
                                         mean(`Student Teacher Ratio`, na.rm = TRUE),
                                           `Student Teacher Ratio`))

```

```{r}

# Calculate average student teacher ratio per county
AvgStudentTeacherRatioCounty = my_data |>
  group_by(`County Name`) |>
  summarise(AvgRatio = mean(`Student Teacher Ratio`, na.rm = TRUE))
AvgStudentTeacherRatioCounty

#add  average student teacher ratio by county to the County Table 
CountyTable = CountyTable |>
  left_join(AvgStudentTeacherRatioCounty, by = c("County" = "County Name"))
```

calculate the aver and median student teacher ratio for each county. Then plot these. Then add reference line for mean or median. then show how far above or below they are.

```{r}
# Calculate a single average and median student teacher ratio use as a refernce line in the plot
OverallAvgStudentTeacherRatio = mean(my_data$`Student Teacher Ratio`, na.rm = TRUE)
OverallMedianStudentTeacherRatio = median(my_data$`Student Teacher Ratio`, na.rm = TRUE)

OverallAvgStudentTeacherRatio
OverallMedianStudentTeacherRatio

#add overall average to County Table
CountyTable$OverallAvg = OverallAvgStudentTeacherRatio

#calculate the difference between the average student teacher ratio for each county and the overall average
CountyTable$DifferenceFromOverallAvg = CountyTable$AvgRatio - CountyTable$OverallAvg

#only keep the first word in the County variable
CountyTable$County = word(CountyTable$County, 1)
```

```{r}
# Create a bar plot of average student teacher ratio by county
base_st = ggplot(CountyTable,
              aes(x=reorder(County, AvgRatio),
                  y=AvgRatio))

base_st = base_st + theme_classic()

plot1 = base_st + 
  geom_col(fill = "gray")

plot1
```

```{r}
titleText='Most counties have a lower student teacher ratio than statewide average'
sub_titleText='Commonwealth of Massachusetts - 2024-2025 School Year'
sourceText='Source: National Center for Education Statistics'

plot2 = plot1 + labs(title=titleText,
                     subtitle=sub_titleText,
                     x =NULL, 
                     y = NULL,
                     caption = sourceText) 

plot2

```

```{r}
#adding the average student teacher ratio for the state as a reference line

plot3 = plot2 + geom_hline( yintercept = 11.95753,
                            linetype = "dotted",
                            linewidth = .7,
                            alpha = 0.8,) +
  scale_y_continuous(breaks=c(0,5,10, 15),
                                 limits = c(0, 15)
                          ) +
  geom_text(aes(label=round(AvgRatio,1)),
            position = position_stack(vjust = 1.05),
            size = 3)

plot3

```

```{r}

base_st = ggplot(CountyTable,
              aes(x=reorder(County, AvgRatio),
                  y=DifferenceFromOverallAvg))

base_st = base_st + theme_classic()
#creating a lollipop plot to show the difference between the average student teacher ratio for each county and the overall average

lolliplot1 = base_st + geom_segment(aes(y=0, 
                                   x= reorder(County, AvgRatio),
                                   yend= DifferenceFromOverallAvg,
                                   xend= reorder(County, AvgRatio)),
                             color="gray", size=0.5) 
lolliplot1
```

```{r}
#adding Postitive/Negative Color
CountyTable$StateAverage=ifelse(CountyTable$DifferenceFromOverallAvg>0,"Higher","Lower") #lower, so negative/no, is good in this case
CountyTable

base_st = ggplot(CountyTable,
              aes(x=reorder(County, AvgRatio),
                  y=DifferenceFromOverallAvg,
                  color=StateAverage,
                  label=round(DifferenceFromOverallAvg,2))
              )

base_st = base_st + theme_classic()

lolliplot1 = base_st + geom_segment(aes(y=0, 
                                   x= reorder(County, AvgRatio),
                                   yend= DifferenceFromOverallAvg,
                                   xend= reorder(County, AvgRatio)),
                             color="gray", size=0.5)

lolliplot2= lolliplot1 + 
  geom_point(size=2) +
  geom_text(nudge_x = .4,
            size=2.4,
            show.legend = FALSE) 
            
#add horizontal line at 0 to show the state average and labels  
lolliplot2 = lolliplot2 + geom_hline(yintercept = 0,
                        linewidth = .7,
                        alpha = 0.8) +
  scale_color_manual(values = c("Higher" = "red", "Lower" = "black")) +
  labs(title="Most counties in Massachusetts have a lower student teacher ratio \nthan the statewide average of 11.96.",
       subtitle="2024-2025 School Year",
       x=NULL,
       y="Difference from State Average Student Teacher Ratio",
       caption="Source: National Center for Education Statistics \n10 records removed for lack of data and 2 outliers imputed with county mean") +
  theme(legend.position = "none")


```

```{r}
#now i will finish the plot getting rid of elements and moving the county label

lolliplot3 = lolliplot2 + theme(axis.ticks.y = element_blank(),
                                axis.title.y = element_blank(),
                                axis.line.y = element_blank(),
                                axis.text.y = element_blank(),
                                axis.ticks.x = element_blank(),
                                axis.title.x = element_blank(),
                                axis.line.x = element_blank()
                                )


lolliplot3 = lolliplot3 + geom_label(aes(label=County),
                                      color ='black ',
                                      size =1.75,
                                      y=0,
                                      show.legend = FALSE ) +
             theme(axis.text.x = element_blank(),
                   plot.title = element_text(face = "bold", size = 10),
                   plot.subtitle = element_text(size = 8),
                   plot.caption = element_text(face = "italic", size = 8)
                   )

lolliplot3

charplot <-lolliplot3
                          
```

```{r}
#add total number of students per county
TotalStudents = my_data |>
  group_by(`County Name`) |>
  summarise(TotalStudents = sum(Students, na.rm = TRUE))

#change county name to only the first word in the TotalStudents table and remove county name variable
TotalStudents$County = word(TotalStudents$`County Name`, 1)
TotalStudents = TotalStudents |>
  select(-`County Name`)

#join total students to the CountyTable by County
CountyTable = CountyTable |>
  left_join(TotalStudents, by = c("County" = "County"))

```

# Numerical Variable

```{r}
#I am going to create a pareto plot showing the total distribution of students across counties


Base_pareto = ggplot(CountyTable,
              aes(x=reorder(County, -TotalStudents),
                  y=TotalStudents))
paretoplot = Base_pareto + geom_col()

paretoplot



```
```{r}

# prepare the data for 'qicharts2'

x <- rep(CountyTable$County, CountyTable$TotalStudents)


paretoplot2 = paretochart(x,x.angle = 45)  

paretoplot2 + 
    labs(title='Cumulative distribution of students across counties \nin Massachusetts',
       subtitle = 'Student Data aggregated and sorted by County',
       caption="Source: National Center for Education Statistics")



```
I cannot really modify the base chart, so i will recreate with ggplot to have more control over the design 
```{r}

# sort data and calculate cumulative percentage for the pareto plot
CountyTable_plot <- CountyTable %>%
  arrange(desc(TotalStudents)) %>%
  mutate(
    County = factor(County, levels = County),
    percent_students = TotalStudents / sum(TotalStudents),
    cumulative_pct = cumsum(TotalStudents) / sum(TotalStudents),
    label = comma(TotalStudents)
  )

```



```{r}
# Create Pareto-style chart 

numplot = ggplot(CountyTable_plot, aes(x = County)) +
  geom_col(aes(y = percent_students), fill = "tan") +
  geom_text(
    aes(y = percent_students, label = label),
    vjust = -0.7,
    size = 1.8
  ) +
  geom_line(
    aes(y = cumulative_pct, group = 1),
    color = "gray",
    linewidth = 1,
    alpha = 0.4
  ) +
  geom_point(
    aes(y = cumulative_pct),
    color = "darkgray",
    size = 2
  ) +
  scale_y_continuous(
    labels = percent,
    limits = c(0, 1.05),
    breaks = seq(0, 1, by = 0.2)
  ) +
  
  labs(
    title = "Cumulative distribution of students across \ncounties in Massachusetts",
    subtitle = "Students aggregated and sorted by county for 2024-2025 school year",
    x = NULL,
    y=NULL,
    caption = "Source: National Center for Education Statistics"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 0.7, size = 6),
    axis.text.y = element_text(size = 6),
    plot.title = element_text(face = "bold", size = 10),
    plot.subtitle = element_text(size = 8),
    plot.caption = element_text(face = "italic", size = 8),
    axis.line.x = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.4),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.x = element_blank()
  )
numplot
```

```{r}
saveRDS(charplot, file = "plot1.rds")
saveRDS(numplot, file = "plot2.rds")
```


