---
title: "hw-3 code"
output: html_notebook
---
First i will load library and clean memory
```{r}
#Load Libraries
rm(list = ls()) # clean memory
library(tidyverse)
library(ggplot2)
```

I will now load my data
```{r}
#Load Data


file_path <- "C:/Users/lgianetti/OneDrive - City of Cambridge/DACSS Program/690V_data_viz/hw-3/ncesdata_MASS_202425.xlsx"

my_data <- read_excel(path = file_path,
    skip = 11)

```

I will do a quick look at the data
```{r}
#Look at Data
str(my_data, width = 70,strict.width='cut')
```
Now I will clean the data by correcting the data types and checking for missing values and outliers. I will also convert the appropriate columns to factors and numeric types as needed.
```{r}
#Clean Data
my_data$`Low Grade` <- factor(my_data$`Low Grade`, ordered = TRUE)
my_data$`High Grade` <- factor(my_data$`High Grade`, ordered = TRUE)
my_data$`Locale Code` <- factor(my_data$`Locale Code`)
my_data$`Locale` <- factor(my_data$`Locale`)
my_data$`Charter` <- factor(my_data$`Charter`)
my_data$`Type` <- factor(my_data$`Type`)
my_data$`Status` <- factor(my_data$`Status`)
my_data$`Students` <- as.numeric(my_data$`Students`)
my_data$`Teachers` <- as.numeric(my_data$`Teachers`)
my_data$`Student Teacher Ratio` <- as.numeric(my_data$`Student Teacher Ratio`)
my_data$`Directly Certified` <- as.numeric(my_data$`Directly Certified`)
my_data$`County Name` <- factor(my_data$`County Name`) 


str(my_data, width = 70,strict.width='cut')
summary(my_data) # check for missing values and outliers


```
Now I will start to explore the data by creating some visualizations. I will create a bar plot to show the distribution of schools by County.

```{r}

# Total values
TotalValues=table(my_data$'County Name',
                 useNA = "ifany") #include all values!
TotalValues

TotalPercent = prop.table(TotalValues)*100
```
```{r}
# Create a data frame for plotting
CountyTable=as.data.frame(TotalValues)

names(CountyTable)=c("County","Count")

CountyTable$Percent=as.vector(TotalPercent)

CountyTable
```
```{r}
# Create a bar plot
base = ggplot(CountyTable,
              aes(x=reorder(County, -Percent),
                  y=Percent))

base + geom_bar(stat="identity")
```

```{r}
#calculate the average student teacher ratio per county and create a bar plot to show the average student teacher ratio by County.
# Calculate average student teacher ratio per county
AvgStudentTeacherRatioCounty = my_data |>
  group_by(`County Name`) |>
  summarise(AvgRatio = mean(`Student Teacher Ratio`, na.rm = TRUE))
AvgStudentTeacherRatioCounty

AvgStudentTeacherRatio = my_data |>
  summarise(AvgRatio = mean(`Student Teacher Ratio`, na.rm = TRUE))
AvgStudentTeacherRatio

MedianStudentTeacherRatio = my_data |>
  summarise(MedRatio = median(`Student Teacher Ratio`, na.rm = TRUE))
MedianStudentTeacherRatio

```
calculate the aver and median student teacher ratio for each county.  Then plot these. Then add reference line for mean or median. then show how far above or below they are.

```{r}
#add student teacher ratio averages to the County Table

CountyTable <-mutate(CountyTable, 
                     AvgSTRatio = AvgStudentTeacherRatioCounty$AvgRatio,
                     MedSTRatio = MedianStudentTeacherRatio$MedRatio,
                     AvgSTDiff = AvgRatio - AvgStudentTeacherRatio$AvgRatio,
                     MedSTDiff = AvgRatio - MedianStudentTeacherRatio$MedRatio
                     )


CountyTable

```
